Tracing prices: A flow-based cost allocation for optimized power systems

### 2020-10-19 paper/flow_based.tex
review some parts, still missing: abstract, graphical abstract, acknowlegdement, review

\documentclass[11pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage[left=1.80cm, right=1.80cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{amsmath,bm}
\usepackage[colorlinks]{hyperref}
\usepackage[backend=biber, bibencoding=utf8]{biblatex}
\usepackage{eurosym}
\usepackage[dvipsnames]{xcolor}
\usepackage{subcaption}
\usepackage{booktabs} % for table (toprule, bottomrule)
\usepackage{enumitem} % for alphabetical enumeration 
\usepackage{accents}
\usepackage[capitalise]{cleveref}

@@

@@

\begin{abstract}
% Maximizing the welfare of all market participants within a power system is a common approach in energy system modelling. It leads to perfectly scheduled operations of generators and ... 
\end{abstract}

\subsection*{Highlights}
\begin{itemize}
 \item Flow Allocation schemes determine peer-to-peer (P2P) allocation between sources and sinks 
 \item P2P allocation can be used to assign payments of consumers to single network assets
 \item Using shadow prices, Capital Expenditures are accurately allolcated to users which profit from investments
 \item The P2P cost assignments are in alignment with the nodal pricing scheme of a welfare-maximized system   
 \item This opens the possibility to transparently integrate transmission constraints and investment costs in national and international development plans 
 \item Applying to a German future scenario shows how higher electricity costs arise for buses which rely on investments in transmission and local backup generation  
\end{itemize}

@@

Today's power systems are subject to a deep and ongoing transformation. The needed shift to from controllable to weather-driven power generation, the constant improvement and innovation of technology require rigorous system planning and international cooperation \cite{pfenninger_energy_2014,schlachtberger_benefits_2017}. The core of the problem manifests in the costs. Firstly, these should be as small as possible while meeting ecological and sociological standards. Secondly, they must be distributed in a fair and transparent manner. Cost-driving aspects or market players should be detected and addressed appropriately. 
In this context, power system modelling has increasingly gained attention during the last years \cite{bazmi_sustainable_2011,pfenninger_energy_2014,pereira_generation_2017}. Many studies show how cost-efficiency and renewable energy are brought together. However, the question of how and, more importantly, upon what grounds costs are distributed was often left open. \\
This study fills this gap. On the basis of an optimized network the operational state of each time step and corresponding characteristics are in detailed considered to allocate all system costs. The presented approach builds upon two basic concepts: Firstly, the revenue of an network asset at the cost-optimum exactly matches its operational and capital expenditures, also known as the zero-profit condition. Secondly, \textit{flow tracing}, following Bialek's Average Participation (AP), assigns the operation of single network assets to consumers in an plausible manner. 
Combining these two relation allows for an transparent allocation of all operational (OPEX) and capital expenditures (CAPEX) to the consumers in the network. \\
Flow-based cost allocation was already discussed and performed in various works \cite{galiana_transmission_2003,shahidehpour_market_2002,meng_investigation_2007,tranberg_flow-based_2018,nikoukar_transmission_2012,arabali_pricing_2012,wu_locational_2005}. Shahidehpour et al. give a deep insight into allocating congestion cost and transmission investments to market participants \cite{shahidehpour_market_2002} with different allocation techniques. Based on Generation Shift Factors, \ie the marginal contribution of generators to a flow on line, the Locational Marginal Price (LMP) can be represented as a superposition of LMP at the reference bus, the price for congestion and a price for losses. The approach in \cite{meng_investigation_2007} expands this relation for contributions based on the AP scheme, which allows for an accurate estimation of the LMP, however does not reflect the exact LMP of an optimized network. A similar approach was taken in \cite{tranberg_flow-based_2018} where electricity prices of a non-optimal power dispatch were allocated using flow tracing.

The following work brings advantages of all the mentioned studies together. It reassures localized cost allocations while fully aligning payments to the nodal pricing scheme based on the LMP. The work serves to facilitate transparency and cost-benefit analysis in network plannings such as the Ten Year Network Development Plan \cite{entso-e_completing_2020} or the German Netzentwicklungsplan (NEP) \cite{bundesnetzagentur_netzentwicklungsplan_2020}. Further, it builds a basis of a usage based transmission cost allocation. The first section which presents the full method splits into: deriving the zero-profit condition for different kind of assets (\cref{sec:zero_profit_rule});  presenting the AP allocation scheme and resulting allocation from assets to consumers (\cref{sec:dispatch_allocations}); presenting the cost allocation and the impact of additional constraints (\cref{sec:general_scheme,sec:design_constraints}); a numerical example (\cref{sec:numerical_example}). The second part in \cref{sec:application_case} applies the cost allocation to a optimized German power system with high shares of renewable power sources and evaluates the allocated costs.

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system is the sum of multiple cost terms $\cost$. Typically, these include operational expenditures for generators $\opexgeneration$, expenditures for emissions $\emissioncost$,  capital investments for the transmission system $\capexflow$ and so on, \ie
\begin{align}
\totalcost &= \sum_\circ \cost = \opexgeneration + \emissioncost +  \capexflow + ...
\label{eq:total_cost}
\end{align}
In turn, each of the terms $\cost$ consists of the costs associated to an asset $i$ in the system, 
\begin{align}
    \cost = \sum_{i} \cost_{i}
    \label{eq:asset_cost}
\end{align} 
where an ``asset'' describes any operating components of the network, such as a generator, line, energy storage etc. We refer to the set of assets as $I$. 
In a long-term equilibrium of a power system with perfect competition, the zero-profit condition states that each cost term $\cost_{i}$ can be considered as a cost-weighted sum of the operational state $s_{i,t}$ of asset $i$ and time $t$, \ie
\begin{align}
    \cost_{i} = \sum_t  \costfactor \, \state
    \label{eq:cost_decomposition}
\end{align}
where $\costfactor$ denotes a corresponding cost factor in \euro/MW. 
If $\cost_i$ describes the OPEX occasioned by asset $i$, the cost factor $\costfactor$ is simply given by the marginal operational price $o_i$. However, as we will show in the following, if it describes the CAPEX of asset $i$, $\costfactor$ is a composition of shadow prices $\mu_{i,t}$ given at the cost-optimum.

For a detailed demonstration, we derive  \cref{eq:cost_decomposition} for each cost term separately while considering generators, transmission lines and storages in the network. Therefore, let $o_i$ denote the operational price per MWh for asset $i$ and $c_i$ the capital price for one MW capacity expansion. \Cref{tab:cost_allocation_map} summarizes all derived relations, these can be inserted into \cref{eq:asset_cost,eq:cost_decomposition} for each single cost term.

@@

\subsubsection*{Generators}
\begin{subequations}
Let $S \subseteq I$ be the set of generators in the network, such that $\generation = s_{s,t}$ describes the power production of generator $s \in S$. The OPEX occasioned by generator $s$ is given by a cost-weighted sum of the production, thus 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
In case a fix price for emissions $\emissionprice$ in \euro\, per tonne-CO$_2$ equivalents, is assumed, a further the cost term per generator $s$     
\begin{align}
 \emissioncost_s = \emissionprice \, \sum_t  \emission \, \generation
\end{align}
adds to $\totalcost$. Here, $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour\, of generator $s$. In contrast to OPEX and emission costs, the CAPEX of $S$ are not a function of the production $\generation$, but of the actual installed capacity $\capacitygeneration$ of generator $s$. In the optimization it limits the generation $\generation$ in the form of 
\begin{align}
\generation - \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}
\end{align}
The constraint yields a shadow-price of $\muuppergeneration$ in literature often denoted as the Quality of Supply. It can be interpreted as the the price per MW that \cref{eq:upper_generation_capacity_constraint} imposes to the system. If $\muuppergeneration$  is bigger than zero, the constraint is binding, which pushes investments in $\capacitygeneration$. As shown in detail in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition}.    %If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary.

% \subsection{CAPEX Allocation}
% For the CAPEX allocation, it becomes crucial to look at the individual relations between operational state $\state$ and its binding constraints.

% \subsubsection{Generators}

% The nominal capacity $\capacitygeneration$ constrains the generation $\generation$ in the form of 
% \begin{align}
% \generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
% \label[constraint]{eq:upper_generation_capacity_constraint}\\ 
% - \generation &\le 0 \resultsin{\mulowergeneration} \Forall{s,t} 
% \label[constraint]{eq:lower_generation_capacity_constraint}
% \end{align}
% where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. At a cost-optimum, these two constraints yield the shadow prices $\muuppergeneration$ and $\mulowergeneration$.  As shown in \cite{brown_decreasing_2020} and in detailed in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
% \begin{align}
%  \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
%  \label{eq:no_profit_capex_generation}
% \end{align}
% This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition}. The resulting the CAPEX allocation is given by
% \begin{align}
%  \allocatecapexgeneration = \muuppergeneration \, \allocategeneration
%  \label{eq:allocate_capexGeneration_detailed}
% \end{align}
% How does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding, which pushes investments in $\capacitygeneration$.  If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
% When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} , we can use \cref{eq:state_allocation_constraint,eq:no_profit_capex_generation} to see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

Let $L \subset I$ be the set of transmission lines in the system, these may include Alternating Current (AC) as well as Directed Current (DC) lines. Further let $\flow = s_{\ell,t}$ represent the power flow on line $\ell \in L$.  If the OPEX of the transmission system is taken into account in $\totalcost$ (these are often neglected in power system models), these may be approximated by $\opexflow_\ell = \sum_t o_\ell |\flow|$, that is, a cost weighted sum of the net flow on line $\ell$. Again this stands in contrast to the CAPEX which not a function of $\flow$ but of the transmission capacity $\capacityflow$. It limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
At the cost-optimum, the two constraints yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the relation that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow (for details see \cref{sec:zero_profit_flow}) 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, the shadow prices $\muupperflow$ and $\mulowerflow$ can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, a non-zero values indicate that \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint} are bound and therefore that the congestion on line $\ell$ at time $t$ is imposing costs to the system.

@@

% The complementary slackness of the above constraints,  
% \begin{align}
%     \capexstorage_r = \capitalpricestorage \, \capacitystorage = \sum_t \muupperstoragedispatch \, \storagedispatch + \muupperstoragecharge \, \storagecharge + \muupperstoragesoc \, \storagesoc
% \end{align}

@@

After representing the cost terms $\cost$ for OPEX and CAPEX by a cost-weighted sum of the operational state $\state$, we now show how the operational state in turn is allocated to consumers. This builds the second crucial ingredient of the cost allocation.

@@

% The Equivalent Bilateral Exchanges (EBE) \cite{galiana_transmission_2003} for example assumes that every producer supplies every consumer proportional to its share in the total consumption. Despite the straightforward principal, this scheme does not suit for large networks where remote buses would interconnect in the same way as buses in close vicinity \cite{gil_multiarea_2005}. The Flow Based Market Coupling (FBMC) is based on predefined zonal sensitivities for allocating intra-regional and inter-regional power flow. However, with its reliance on the mentioned sensitivity, the Generation Shift Keys, the scheme is strongly dependent on exogenous assumptions.

@@

Yet, we didn't touch the allocation from transporting assets $i \in L$ to consumers. To tackle this, let $\ptdf$ denote a linear mapping between the injection $\left(\nodalgeneration[m] - \demand[m]\right)$ and the flow $\flow$, such that  
\begin{align}
 \flow  = \sum_m \ptdf[m]\, \left(\nodalgeneration[m] - \demand[m]\right) \Forall{\ell \in L, t}
 \label{eq:power_flow_equation}  
\end{align}
Usually, $\ptdf$ is given by the Power Transfer Distribution Factors (PTDF) which indicate the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $m$. For transport models or networks with High Voltage Directed Current (HVDC) lines, these can calculated or extended using the formulation presented in \cite{hofmann_flow_2020-1}.
Inserting \cref{eq:nodalgeneration} into \cref{eq:power_flow_equation} and expanding the sum yields 
\begin{align}
    \allocateflow = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n m} \demand \right) \Forall{\ell \in L, n, t}
 \label{eq:allocate_flow}
\end{align}
Complementary to \cref{eq:allocate_production}, this allocation indicates the power flow on line $\ell$ and time $t$ which is ending up at node $n$ and therefore consumed by $\demand$.  
Note that the assignment of line flows to consumers is different from the traced flow returned by the AP scheme. As pointed out in \cite{galiana_transmission_2003} the AP scheme, does not obey the Kirchhoff Voltage Law but only the Kirchhoff Current Law. As we show later and in \cref{sec:proof_equivalence} a consideration of both is needed in order to align the allocated costs with the Locational Marginal Prices.

@@

Using the presented relations, we are able to straightforwardly define the full cost allocation. Therefore, the expression in \cref{eq:decompose_state} is inserted in \cref{eq:cost_decomposition} which leads to 
\begin{align}
    \allocatecost = \costfactor \, \allocatestate
    \label{eq:cost_allocation}
\end{align}
By default, all cost of a cost term $\cost_i$ associated with asset $i$ are allocated, thus $\cost_i = \sum_n \allocatecost$; likewise all system costs are distributed, $\totalcost = \sum_{\circ, i, n, t} \allocatecost$.

The cost allocation entails a further important property. When summing over all assets $i$ and cost terms $\circ$, the cost allocation yields $\payment$, which we refer to as the nodal payment or payment. This exactly matches the payment determined by the Locational Marginal Price (LMP) $\lmp$, thus 
\begin{align}
    \payment = \sum_{\circ, i} \allocatecost  = \lmp\, \demand  \Forall{n,t}
    \label{eq:payment}
\end{align}
In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the change of costs for an incremental increase of electricity demand $\demand$ at node $n$ and time $t$ \cite{schweppe_spot_1988}. Mathematically this translates to the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$, $\lmp = \partial \totalcost /\partial \demand$. At the optimum the LMP is given by a shadow price of the nodal balance constraint (see \cref{sec:lmp} for details). 
\Cref{eq:payment} which is in detailed proved in \cref{sec:proof_equivalence} connects the cost allocation to a nodal pricing scheme implied by the LMP.

@@

% \subsection{Connection to Locational Marginal Prices}

% % 
% \begin{figure}[h]
% \centering
% \includegraphics[width=.8\linewidth]{price_decomposition.png}
% \caption{Schematic decomposition of the Locational Market Price $\lmp$. As the cost function of the power system model different cost terms as indicated by \cref{eq:total_cost}, so does each $\lmp$ at node $n$ and time $t$.}
% \label{fig:price_decomposition}
% \end{figure}

@@

% % 
% This leads to a nodal pricing where over the span of optimized time steps $t$, the system costs are partially or totally payed back by the consumers 
% \begin{align}
% \totalcost - \remainingcost &=  \sum_{n,t} \lmp \, \demand
% \label{eq:total_revenue}
% \end{align}
% depending on the costs $\remainingcost$ which are independent of the nodal demand  
% \begin{align}
%  \pdv{\remainingcost}{\demand} = 0
% \end{align}
% % 
% Generally speaking, the cost term $\remainingcost$, not covered by the consumers, results from additional request to the network design, such as capacity expansion limits or minimum share of one technology in the power mix. However, in most cases, where $\remainingcost \ll  \totalcost$, these play a minor role.

% However the question of how the LMP can be decomposed into contributions of single cost terms $\cost_{i}$ associated with asset $i$ remains unanswered. This work aims at presenting and illustrating a an intuitive, peer-to-peer cost allocation including all network assets.

@@

Power system modelling does rarely follow a pure Greenfield approach with unlimited capacity expansion. Rather, today's models are setting various constraints defining socio-political or  technical requirements. However, this will alter the equality of total cost and total revenue. More precisely, each constraint $h_j$ (other then the nodal balance constraint) of the form 
\begin{align}
    h_j \left(\state, \capacity \right) < K
    \label{eq:design_constraint}
\end{align}
where $K$ is any non-zero constant and $\capacity$ denotes the nominal capacities of asset $i$, will  alter \cref{eq:cost_decomposition} to 
\begin{align}
    \cost_i - \remainingcost^\circ_i &= \sum_t \costfactor \, \state \\
    &= \sum_{n,t} \allocatecost
\end{align}
% We define the allocated costs associated to cost term $\cost$ is then given by 
% \begin{align}
%     \bar{\cost} = \sum_i  \cost_i - \remainingcost_i
%     \label{altered_cost}
% \end{align}
According to the nature of \cref{eq:design_constraint} and the corresponding $\remainingcost_i$, it is either larger, equal or lower than $\cost$. Whereas \cref{eq:payment} still holds true, the total of payments do not return the total system cost $\totalcost$ anymore. The total mismatch $\remainingcost$ is given by 
\begin{align}
    \remainingcost = \totalcost - \sum_{n,t} \payment
\end{align}
In the following we highlight two often used classes of designs constraints in the form of \cref{eq:design_constraint} and show how to incorporate them into the cost allocation.

@@

\section{Limitations and possible Extensions}

>>>>

\documentclass[11pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage[left=1.80cm, right=1.80cm, top=2.00cm, bottom=2.00cm]{geometry}
\usepackage{amsmath,bm}
\usepackage[colorlinks]{hyperref}
\usepackage[natbib=true, sorting=none, backend=biber, bibencoding=utf8]{biblatex}
\usepackage{eurosym}
\usepackage[dvipsnames]{xcolor}
\usepackage{subcaption}
\usepackage{booktabs} % for table (toprule, bottomrule)
\usepackage{enumitem} % for alphabetical enumeration 
\usepackage{accents}
\usepackage[capitalise]{cleveref}

@@

\twocolumn[
  \begin{@twocolumnfalse}

@@

    \begin{abstract}
        Maximizing the welfare of all market participants within a power system is a common approach in energy system modelling. It leads to perfectly scheduled operations and investments while securing all power demands. \dots 
        This work introduces a peer-to-peer cost allocation for welfare-maximized systems   \dots
        This opens the possibility to transparently integrate transmission constraints and investment costs in national and international development plans.
    \end{abstract}

\end{@twocolumnfalse}
]

\subsection*{Highlights}
\begin{itemize}
    \item In long-term investment models every network asset gains back its variable and investment cost from its operation based revenue
    \item Flow tracing is used as a basis to allocate the operation of assets to consumers. 
    \item Using operational and shadow prices, the allocations assign all system costs in an P2P manner.
    \item The P2P cost assignments are locally constraint while being in alignment with the nodal pricing scheme of a welfare-maximized system.    
    \item Applying to a German future scenario shows how higher electricity costs arise for buses relying on investments in transmission and local backup generation  
\end{itemize}

@@

Today's power systems are subject to a deep and ongoing transformation. The needed shift to from controllable to weather-driven power generation, the constant improvement and innovation of technology require rigorous system planning and international cooperation \cite{pfenninger_energy_2014,schlachtberger_benefits_2017}. The core of the problem manifests in the costs. Firstly, these should be as small as possible while meeting ecological and sociological standards. Secondly, they must be distributed in a fair and transparent manner. Cost-driving aspects or market players should be detected and addressed appropriately. 
In this context, power system modelling has increasingly gained attention during the last years \cite{pfenninger_energy_2014,bazmi_sustainable_2011,pereira_generation_2017}. Many studies show how cost-efficiency and renewable energy are brought together. However, the question of how and, more importantly, upon what grounds costs are distributed was often left open. \\
This study bridges this gap. On the basis of an optimized network the operational state of each time step and corresponding characteristics are in detailed considered to allocate all system costs. The presented approach builds upon two basic concepts: Firstly, the revenue of an network asset at the cost-optimum exactly matches its operational and capital expenditures, also known as the zero-profit condition. Secondly, flow tracing, following Bialek's Average Participation (AP) \cite{bialek_tracing_1996}, can be used to assigns the operation of single network assets to consumers in an plausible and locally constraint manner. 
Combining these two relations allows for an transparent allocation of all operational (OPEX) and capital expenditures (CAPEX) to the consumers in the network. \\
Flow-based cost allocation was already discussed and performed in various works \cite{galiana_transmission_2003,shahidehpour_market_2002,meng_investigation_2007,tranberg_flow-based_2018,nikoukar_transmission_2012,arabali_pricing_2012,wu_locational_2005}. Shahidehpour et al. give a deep insight into allocating congestion cost and transmission investments to market participants \cite{shahidehpour_market_2002} with different allocation techniques. Based on Generation Shift Factors, \ie the marginal contribution of generators to a flow on line, the Locational Marginal Price (LMP) can be represented as a superposition of LMP at the reference bus, the price for congestion and a price for losses. The approach in \cite{meng_investigation_2007} expands this relation for contributions based on the AP scheme, which allows for an accurate estimation of the LMP, however does not reflect the exact LMP of an optimized network. A similar approach was taken in \cite{tranberg_flow-based_2018} where electricity prices of a non-optimal power dispatch were allocated using flow tracing.

The following work brings advantages of all the mentioned studies together. It reassures localized cost allocations while fully aligning payments to the nodal pricing scheme based on the LMP. The work serves to facilitate transparency and cost-benefit analysis in network plannings such as the Ten Year Network Development Plan \cite{entso-e_completing_2020} or the German Netzentwicklungsplan (NEP) \cite{bundesnetzagentur_netzentwicklungsplan_2020}. Further, it builds a basis of a usage based transmission cost allocation.

The first section which presents the full method splits into the presentation of: the zero-profit condition for different kind of assets (\cref{sec:zero_profit_rule});  the AP allocation scheme and resulting allocation from assets to consumers (\cref{sec:dispatch_allocations}); the cost allocation and the impact of additional constraints (\cref{sec:general_scheme,sec:design_constraints}); a numerical example (\cref{sec:numerical_example}). The second part in \cref{sec:application_case} applies the cost allocation to a optimized German power system with high shares of renewable power sources and evaluates the allocated costs.

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system is the sum of multiple cost terms $\cost$. Typically, these include operational expenditures for generators $\opexgeneration$, expenditures for emissions $\emissioncost$,  capital investments for the transmission system $\capexflow$ and so on, \ie
\begin{align}
\totalcost &= \sum_\circ \cost = \opexgeneration + \emissioncost +  \capexflow + ...
\label{eq:total_cost}
\end{align}
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{cost_decomposition.png}
\caption{Schematic decomposition of the total system cost $\totalcost$ in an long-term investment model.}
\label{fig:cost_decomposition}
\end{figure}
In turn, each of the terms $\cost$ consists of the costs associated to an asset $i$ in the system, 
\begin{align}
    \cost = \sum_{i} \cost_{i}
    \label{eq:asset_cost}
\end{align} 
where an ``asset'' describes any operating components of the network, such as a generator, line, energy storage etc. We refer to the set of assets as $I$. 
In a long-term equilibrium of a power system with perfect competition, the zero-profit condition states that each cost term $\cost_{i}$ can be considered as a cost-weighted sum of the operational state $s_{i,t}$ of asset $i$ and time $t$, \ie
\begin{align}
    \cost_{i} = \sum_t  \costfactor \, \state
    \label{eq:cost_decomposition}
\end{align}
where $\costfactor$ denotes a corresponding cost factor in \euro/MW. 
If $\cost_i$ describes the OPEX occasioned by asset $i$, the cost factor $\costfactor$ is simply given by the marginal operational price $o_i$. However, as we will show in the following, if it describes the CAPEX of asset $i$, $\costfactor$ is a composition of shadow prices $\mu_{i,t}$ given from the corresponding constraint at the cost-optimum.

For a detailed demonstration, we derive  \cref{eq:cost_decomposition} for generators, transmission lines and storages separately. Therefore, let $o_i$ denote the operational price per MWh of asset $i$ and $c_i$ the capital price for one MW capacity expansion. \Cref{tab:cost_allocation_map} summarizes all derived relations, these can be inserted into \cref{eq:asset_cost,eq:cost_decomposition} for each single cost term.

@@

\subsubsection*{Generators}
\begin{subequations}
Let $S \subseteq I$ be the set of generators in the network, such that $\generation = s_{s,t}$ describes the power production of generator $s \in S$. The OPEX occasioned by generator $s$ is given by a cost-weighted sum of the production, thus 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
In case a fix price for emissions $\emissionprice$ in \euro\, per tonne-CO$_2$ equivalents, is assumed, a further the cost term per generator $s$     
\begin{align}
 \emissioncost_s = \emissionprice \, \sum_t  \emission \, \generation
\end{align}
adds to $\totalcost$. Here, $\emission$ denotes the emission factor in tonne-CO$_2$ per \megawatthour\, of generator $s$. In contrast to OPEX and emission costs, the CAPEX of $S$ are not a function of the production $\generation$, but of the actual installed capacity $\capacitygeneration$ of generator $s$. In the optimization it limits the generation $\generation$ in the form of 
\begin{align}
\generation - \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}
\end{align}
The constraint yields a shadow-price of $\muuppergeneration$ in literature often denoted as the Quality of Supply \cite{schweppe_spot_1988}. It can be interpreted as the price per MW that \cref{eq:upper_generation_capacity_constraint} imposes to the system. If $\muuppergeneration$  is bigger than zero, the constraint is binding, which pushes investments in $\capacitygeneration$. As shown in \cite{brown_decreasing_2020} and in detail in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition}.   
\end{subequations}

@@

Let $L \subset I$ be the set of transmission lines in the system, these may include Alternating Current (AC) as well as Directed Current (DC) lines. Further let $\flow = s_{\ell,t}$ represent the power flow on line $\ell \in L$.  If the OPEX of the transmission system is taken into account in $\totalcost$ (these are often neglected in power system models), these may be approximated by $\opexflow_\ell = \sum_t o_\ell |\flow|$, that is, a cost weighted sum of the net flow on line $\ell$. Again this stands in contrast to the CAPEX which not a function of $\flow$ but of the transmission capacity $\capacityflow$. It limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
At the cost-optimum, the two constraints yield the shadow prices $\muupperflow$ and $\mulowerflow$. We use the relation that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow (for details see \cref{sec:zero_profit_flow}) 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, the shadow prices $\muupperflow$ and $\mulowerflow$ can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, a non-zero values indicate that \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint} are bound and therefore that the congestion on line $\ell$ at time $t$ is imposing costs to the system.

@@

@@

The fact that all asset related costs $\cost_i$ can be represented as a cost-weighted sum of the operational state $\state$, prompts the question how the operational state $\state$, in turn, is allocated to consumers. This builds the second crucial ingredient of the cost allocation.

@@

@@

Yet, we didn't touch the allocation of transporting assets $i \in L$ to consumers. Note that the traced flow based in the AP scheme does obey the Kirchhoff Current Law but not the Kirchhoff Voltage Law, as already pointed out in \cite{galiana_transmission_2003}. However, as we show later a consideration of both laws is needed in order to align the allocated costs with the optimized Locational Marginal Prices. To tackle this, let $\ptdf$ denote a linear mapping between the injection $\left(\nodalgeneration[m] - \demand[m]\right)$ and the flow $\flow$, such that  
\begin{align}
 \flow  = \sum_m \ptdf[m]\, \left(\nodalgeneration[m] - \demand[m]\right) \Forall{\ell \in L, t}
 \label{eq:power_flow_equation}  
\end{align}
Usually, $\ptdf$ is given by the Power Transfer Distribution Factors (PTDF) which indicate the changes in the flow on line $\ell$ for one unit (typically one MW) of net power production at bus $m$. For transport models or networks with High Voltage Directed Current (HVDC) lines, these can retrospectively be calculated or extended using the formulation presented in \cite{hofmann_flow_2020-1}.
Inserting \cref{eq:nodalgeneration} into \cref{eq:power_flow_equation} and expanding the sum yields 
\begin{align}
    \allocateflow = \sum_m \ptdf[m] \left( \allocatepeer  - \delta_{n m} \demand \right) \Forall{\ell \in L, n, t}
 \label{eq:allocate_flow}
\end{align}
Complementary to \cref{eq:allocate_production}, this allocation indicates the power flow on line $\ell$ and time $t$ which is finally consumed by $\demand$.

@@

Using the presented relations, we are able to straightforwardly define the full cost allocation. Therefore, the expanded sum in \cref{eq:decompose_state} is inserted in \cref{eq:cost_decomposition} which leads to 
\begin{align}
    \allocatecost = \costfactor \, \allocatestate
    \label{eq:cost_allocation}
\end{align}
By default, all cost of a cost term $\cost_i$ associated with asset $i$ are allocated, thus $\cost_i = \sum_n \allocatecost$; likewise all system costs are distributed, $\totalcost = \sum_{\circ, i, n, t} \allocatecost$.

The cost allocation entails a further important property. In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the change of costs for an incremental increase of electricity demand $\demand$ at node $n$ and time $t$ \cite{schweppe_spot_1988}. Mathematically this translates to the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$, $\lmp = \partial \totalcost /\partial \demand$. At the optimum this quantity is given by a shadow price of the nodal balance constraint (see \cref{sec:lmp} for details). Now, when summing over all assets $i$ and cost terms $\circ$, the cost allocation yields 
\begin{align}
    \payment = \sum_{\circ, i} \allocatecost  = \lmp\, \demand  \Forall{n,t}
    \label{eq:payment}
\end{align} 
which we refer to as the nodal payment or payment. It exactly matches the payment determined by the LMP $\lmp$. Hence, \Cref{eq:payment} which is in detailed proved in \cref{sec:proof_equivalence} shows that the cost allocation is embedded in the nodal pricing scheme implied by the LMP.

@@

% \begin{figure}[h]
% \centering
% \includegraphics[width=.8\linewidth]{price_decomposition.png}
% \caption{Schematic decomposition of the Locational Market Price $\lmp$. As the cost function of the power system model different cost terms as indicated by \cref{eq:total_cost}, so does each $\lmp$ at node $n$ and time $t$.}
% \label{fig:price_decomposition}
% \end{figure}

@@

@@

Power system modelling does rarely follow a pure Greenfield approach with unlimited capacity expansion. Rather, today's models are setting various constraints defining socio-political or  technical requirements. However, this will alter the equality of total cost and total revenue. More precisely, each constraint $h_j$ (other then the nodal balance constraint) of the form 
\begin{align}
    h_j \left(\state, \capacity \right) < K
    \label{eq:design_constraint}
\end{align}
where $K$ is any non-zero constant and $\capacity$ denotes the nominal capacities of asset $i$, will  alter \cref{eq:cost_decomposition} to 
\begin{align}
    \cost_i - \remainingcost^\circ_i &= \sum_t \costfactor \, \state \\
    &= \sum_{n,t} \allocatecost
\end{align}

According to the nature of \cref{eq:design_constraint} and the corresponding $\remainingcost_i$, it is either larger, equal or lower than $\cost$. Whereas \cref{eq:payment} still holds true, the total of payments do not return the total system cost $\totalcost$ anymore. The total mismatch $\remainingcost$ is given by 
\begin{align}
    \remainingcost = \totalcost - \sum_{n,t} \payment
\end{align}
In the following we highlight two often used classes of designs constraints in the form of \cref{eq:design_constraint} and show how to consider them into the cost allocation.

@@

\section{Conclusion and Outlook}
